---
title: "life expectancy"
author: "Robert de Ridder"
date: "2026-09-18"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Abstract

This study investigates whether life expectancy can be predicted using socio-economic 
indicators from Our World in Data. Several predictive models were developed, 
including linear regression, K-nearest neighbours, and random forests. 
Model performance was evaluated using RMSE. The random forest model achieved 
the best performance with a final test RMSE of 1.8 years. Fertility rate and 
healthcare expenditure emerged as the most influential predictors. GDP per 
capita was strongly correlated with life expectancy but added limited predictive 
value once other socio-economic indicators were included.

# Introduction

The course HarvardX PH125.9x: Data Science Capstone is the ninth and final 
component of the Data Science Professional Certificate program. Within this course, 
students are required to complete two projects that together determine the final grade.
This report describes the second project, in which students are free to select 
their own research topic. For this study, the chosen topic is an analysis of 
socio-economic factors that may influence life expectancy at birth.

The central research questions are:

1.	To what extent can life expectancy at birth be predicted using socio-economic indicators?
2.	Which socio-economic factors contribute most strongly to these predictions?
3.	Which modelling approach provides the best predictive performance?

Life expectancy at birth represents the average number of years a newborn is 
expected to live. Over the past decades, substantial improvements in public health 
have been achieved worldwide. According to the World Health Organization (WHO), 
global life expectancy increased significantly between 2000 and 2019. However, 
this overall progress masks considerable differences between countries and 
population groups. Individuals with a lower socio-economic status often have less 
access to healthcare and education, which may negatively affect their health 
outcomes and life expectancy (World Health Organization, n.d.).
In addition to access to healthcare and education, several other factors may 
contribute to differences in life expectancy. This study utilizes data obtained 
from Our World in Data (OWID) (Our World in Data, n.d.). The dataset contains a wide range of indicators 
related to economic conditions, health, lifestyle, and demographics.

The following variables are included in the analysis:

-	Healthcare spending (% of GDP)
-	Education spending (% of GDP)
-	Years of schooling (average years of education)
-	Vaccination coverage (multiple vaccination indicators)
-	Obesity prevalence (share of adults classified as obese)
-	Fertility rate (children per woman)
-	Urbanization (urban and rural population shares)
-	Renewable energy (share of total energy consumption from renewable sources)
-	Democracy index (ranging from 0 to 1, where 1 represents the highest level of democracy)
-	GDP per capita (in U.S. dollars)

The target variable in this study is life expectancy at birth. Using various data analysis 
techniques and predictive models, the study investigates which factors are most strongly associated
with life expectancy and to what extent life expectancy can be predicted accurately.

# Approach

The final dataset was constructed by combining eleven separate datasets 
obtained from Our World in Data (OWID). From each dataset, only the variables 
relevant to this study were selected. Subsequently, all datasets were transformed 
into a common structure, allowing them to be merged into a single integrated dataset.
The healthcare dataset contained the smallest number of observations and was 
therefore used as the foundation for the merging process. This approach was 
chosen to minimize the number of missing values introduced during data integration. 
Despite this strategy, the combined dataset still contained a substantial amount 
of missing data.

```{r create_dataset, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
################################################################################
# Load library
################################################################################

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(ggplot2)) install.packages("ggplot2", repos = "http://cran.us.r-project.org")
if(!require(ggthemes)) install.packages("ggthemes", repos = "http://cran.us.r-project.org")
if(!require(knitr)) install.packages("knitr", repos = "http://cran.us.r-project.org")
if(!require(kableExtra)) install.packages("kableExtra", repos = "http://cran.us.r-project.org")
if(!require(lubridate)) install.packages("lubridate", repos = "http://cran.us.r-project.org")
if(!require(stringr)) install.packages("stringr", repos = "http://cran.us.r-project.org")
if(!require(readxl)) install.packages("readxl", repos = "http://cran.us.r-project.org")
if(!require(randomForest)) install.packages("randomForest", repos = "http://cran.us.r-project.org")
if(!require(zoo)) install.packages("zoo", repos = "http://cran.us.r-project.org")
if(!require(naniar)) install.packages("naniar", repos = "http://cran.us.r-project.org")
if(!require(corrplot)) install.packages("corrplot", repos = "http://cran.us.r-project.org")
if(!require(missForestPredict)) install.packages("missForestPredict", repos = "http://cran.us.r-project.org")
if(!require(broom)) install.packages("broom", repos = "http://cran.us.r-project.org")
if(!require(knitr)) install.packages("knitr", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(caret)
library(ggplot2)
library(ggthemes)
library(knitr)
library(kableExtra)
library(lubridate)
library(stringr)
library(readxl)
library(randomForest)
library(zoo)
library(naniar)
library(corrplot)
library(missForestPredict)
library(broom)
library(knitr)

################################################################################
# Load data
################################################################################

# Note: this process could take a couple of minutes

options(timeout = 500)

# -----------------------
# Our World in Data URLs
# -----------------------

life_url <-
  "https://ourworldindata.org/grapher/life-expectancy-at-different-ages.csv?v=1&csvType=full&useColumnShortNames=true"

gdp_url <-
  "https://ourworldindata.org/grapher/gdp-per-capita-penn-world-table.csv?v=1&csvType=full&useColumnShortNames=true"

healthcare_url <-
  "https://ourworldindata.org/grapher/public-health-expenditure-share-gdp.csv?v=1&csvType=full&useColumnShortNames=true"

schooling_url <-
  "https://ourworldindata.org/grapher/years-of-schooling.csv?v=1&csvType=full&useColumnShortNames=true&metric_type=average_years_schooling&level=all&sex=both"

vaccination_url <-
  "https://ourworldindata.org/grapher/vaccination-coverage-who-unicef.csv?v=1&csvType=full&useColumnShortNames=true&metric=coverage&antigen=comparison"

education_spending_url <-
  "https://ourworldindata.org/grapher/education-spending.csv?v=1&csvType=full&useColumnShortNames=true&spending_type=gdp_share&level=all"

obesity_url <-
  "https://ourworldindata.org/grapher/share-of-adults-defined-as-obese.csv?v=1&csvType=full&useColumnShortNames=true"

fertility_url <-
  "https://ourworldindata.org/grapher/children-born-per-woman.csv?v=1&csvType=full&useColumnShortNames=true"

urbanization_url <-
  "https://ourworldindata.org/grapher/share-of-people-living-in-urban-and-rural-areas-degurba.csv?v=1&csvType=full&useColumnShortNames=true"

renewables_url <-
  "https://ourworldindata.org/grapher/energy-mix.csv?v=1&csvType=full&useColumnShortNames=true&source=renewables&metric=share"

democracy_url <-
  "https://ourworldindata.org/explorers/democracy.csv?v=1&csvType=full&useColumnShortNames=true&Dataset=Varieties+of+Democracy&Metric=Electoral+democracy&Sub-metric=Main+index"

# -----------------------
# Import datasets
# -----------------------

life_at_birth <- read.csv(life_url)                    # life expectancy
gdp_per_capita <- read.csv(gdp_url)                    # GDP per capita ($)
spending_on_education <- read.csv(education_spending_url) # share of gdp
healthcare_spending <- read.csv(healthcare_url)        # healthcare spending (% GDP)
years_of_schooling <- read.csv(schooling_url)          # average years of schooling
vaccination_coverage <- read.csv(vaccination_url)      # vaccination coverage
share_of_adults_defined_as_obese <- read.csv(obesity_url)
fertility_rate <- read.csv(fertility_url)              # children per woman
urbanization <- read.csv(urbanization_url)             # urban/rural population
renewables <- read.csv(renewables_url)                 # renewable energy share
democracy <- read.csv(democracy_url)                   # 0 to 1. 1 = most demacratic

################################################################################
# select columns of interest and rename
################################################################################

life_at_birth <- life_at_birth |> select(entity, year, life_expectancy_0) |>
  rename(life_expectancy = life_expectancy_0)
#############################################
gdp_per_capita <- gdp_per_capita |> select(entity, year, rgdpo_pc) |> 
  rename( gdp_pc = rgdpo_pc)
#############################################
spending_on_education <- spending_on_education |> 
  select(entity, year, combined_expenditure_share_gdp) |>
  rename(spending_edu = combined_expenditure_share_gdp)
#############################################
healthcare_spending <- healthcare_spending |> 
  select(entity, year, share_gdp) |>
  rename(spending_healthcare = share_gdp)
#############################################
years_of_schooling <- years_of_schooling |>
  select(entity, year, mys__sex_total) |>
  rename(years_in_school = mys__sex_total)
#############################################
vaccination_coverage <- vaccination_coverage |>
  select(-code)
#############################################
share_of_adults_defined_as_obese <- share_of_adults_defined_as_obese |>
  select(entity, year, obesity_among_adults__bmi__gt__30_kg_m2__crude_estimate__pct__sex_both_sexes__age_group_18plus__years_of_age) |>
  rename(share_obesity = obesity_among_adults__bmi__gt__30_kg_m2__crude_estimate__pct__sex_both_sexes__age_group_18plus__years_of_age)
#############################################
fertility_rate <- fertility_rate |>
  select(entity, year, fertility_rate_hist) |>
  rename(children_per_woman = fertility_rate_hist)
#############################################
urbanization <- urbanization |> 
  select(entity, year, value__metric_popshare__location_type_rural_total__data_type_estimates,
         value__metric_popshare__location_type_urban_total__data_type_estimates) |>
  rename(move_from_city = value__metric_popshare__location_type_rural_total__data_type_estimates,
         move_to_city = value__metric_popshare__location_type_urban_total__data_type_estimates)

urbanization <- urbanization |> # take all years below 2025 and fill the caps between the 5 year interval
  select(entity, year, move_from_city,
         move_to_city) |>
  filter(year < 2025)

urbanization <- urbanization %>%
  arrange(entity, year) %>%
  group_by(entity) %>%
  complete(year = min(year):max(year)) %>%
  mutate(
    move_from_city =
      na.approx(
        move_from_city,
        x = year,
        na.rm = FALSE
      ),
    move_to_city =
      na.approx(
        move_to_city,
        x = year,
        na.rm = FALSE
      )
  )
#############################################
renewables <- renewables |>
  select(entity, year, renewables_share_pct)
#############################################
democracy <- democracy |>
  select(entity, year, electdem_vdem__estimate_best) |>
  rename(elect_dem_index = electdem_vdem__estimate_best)

################################################################################
# create new dataset
################################################################################

raw_data <- healthcare_spending %>% # join by healthcare to avoid alot of na"s
  left_join(life_at_birth, by = c("entity", "year")) %>%
  left_join(gdp_per_capita, by = c("entity", "year")) |>
  left_join(spending_on_education, by = c("entity", "year")) |>
  left_join(years_of_schooling, by = c("entity", "year")) |>
  left_join(vaccination_coverage, by = c("entity", "year")) |>
  left_join(share_of_adults_defined_as_obese, by = c("entity", "year")) |>
  left_join(fertility_rate, by = c("entity", "year")) |>
  left_join(urbanization, by = c("entity", "year")) |>
  left_join(renewables, by = c("entity", "year")) |>
  left_join(democracy, by = c("entity", "year"))

# create table with missing values

na_summary <- sort(colMeans(is.na(raw_data)), decreasing = TRUE)

na_table <- data.frame(
  Variable = names(na_summary),
  Missing_Percentage = round(na_summary * 100, 1)
) %>%
  filter(Missing_Percentage > 0)

kable(
  na_table,
  caption = "Variables with missing values"
)
```

Variables with more than 50% missing values were removed from the analysis. 
The remaining missing values were imputed using the missForest algorithm after 
the data had been divided into training, validation, and final test sets. 

This approach was selected because previous research has shown that missForest 
can effectively handle different data types and often outperforms several traditional 
imputation methods. An additional advantage is that the method provides an 
out-of-bag estimate of the imputation error (Stekhoven and Bühlmann, 2012).
Furthermore, the relationships among variables in this dataset are likely to be 
both linear and non-linear, making simple median or mean imputation less appropriate, 
as these methods may fail to preserve the underlying structure of the data.

An 85% / 15% split was used to obtain a sufficiently large independent final test set,
while keeping enough observations for model training. The remaining 85% of the data
was further divided into 80% training and 20% validation data. This resulted in 
approximately 68% of the original data being used for training, 17% for validation, 
and 15% for final testing.

To avoid data leakage, the imputation model was fitted exclusively on the training 
data and then applied to the validation and final test sets.

```{r missing_values, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
################################################################################
# remove columns with more than 50% na's
################################################################################

raw_data <- raw_data |>
  select(-coverage__antigen_rotac, -coverage__antigen_ipv1, -coverage__antigen_pcv3,
         -coverage__antigen_hepb3)

raw_data <- raw_data[!is.na(raw_data$life_expectancy), ] # remove na's out of target

################################################################################
# split data in work and final test set
################################################################################

set.seed(123)

train_index <- createDataPartition(raw_data$life_expectancy,
                                   p = 0.85,
                                   list = FALSE)

work_set <- raw_data[train_index, ]
final_test_set  <- raw_data[-train_index, ]

################################################################################
# split work_set in train and validation
################################################################################

set.seed(123)

work_index <- createDataPartition(work_set$life_expectancy,
                                   p = 0.8,
                                   list = FALSE)

train <- work_set[work_index, ]
validate  <- work_set[-work_index, ]


################################################################################
# deal with missing values on train, test and final test set
################################################################################

################################################################################
# Remove identifiers
################################################################################

train <- train |> 
  select(-entity, -year)

validate <- validate |> 
  select(-entity, -year)

final_test_set <- final_test_set |> 
  select(-entity, -year)


################################################################################
# imputation
# Fit preprocessing only on training data
################################################################################

# Separate target from predictors
x_train <- train |> 
  select(-life_expectancy)

x_validate <- validate |> 
  select(-life_expectancy)

x_final_test <- final_test_set |> 
  select(-life_expectancy)


# Fit imputation model on training data only
set.seed(123)

imputation_model <- missForestPredict::missForest(
  x_train,
  maxiter = 10,
  initialization = "median/mode",
  save_models = TRUE,
  num.threads = 2,
  verbose = FALSE
)


# Apply the same preprocessing to all partitions
x_train_imp <- imputation_model$ximp

x_validate_imp <- missForestPredict::missForestPredict(
  imputation_model,
  newdata = x_validate
)

x_final_test_imp <- missForestPredict::missForestPredict(
  imputation_model,
  newdata = x_final_test
)


################################################################################
# Recombine target and predictors
################################################################################

train <- cbind(
  life_expectancy = train$life_expectancy,
  x_train_imp
)

validate <- cbind(
  life_expectancy = validate$life_expectancy,
  x_validate_imp
)

final_test_set <- cbind(
  life_expectancy = final_test_set$life_expectancy,
  x_final_test_imp
)



```

Following imputation and data preprocessing, no missing values remained 
in the training, validation, or final test sets.

```{r missing_value, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
################################################################################
# Check missing values
################################################################################

if (
  sum(is.na(train)) == 0 &
  sum(is.na(validate)) == 0 &
  sum(is.na(final_test_set)) == 0
) {
  cat("No missing values were found in the training, validation, or final test set.\n")
}
```

To assess whether the data partitions are representative of the original dataset, 
a boxplot was constructed. It shows that the distributions of life expectancy 
are similar across the training, validation, and final test sets, 
suggesting that the data split was successful.

```{r boxplot, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

################################################################################
# Check train, validate and final test set data
################################################################################

# Check if train and valid data are representative

boxplot(
  train$life_expectancy,
  validate$life_expectancy,
  final_test_set$life_expectancy
)

# create table with stats

summary_table <- data.frame(
  Dataset = c("Train", "Validation", "Test"),
  N = c(nrow(train), nrow(validate), nrow(final_test_set)),
  Mean = c(
    mean(train$life_expectancy),
    mean(validate$life_expectancy),
    mean(final_test_set$life_expectancy)
  ),
  SD = c(
    sd(train$life_expectancy),
    sd(validate$life_expectancy),
    sd(final_test_set$life_expectancy)
  ),
  Min = c(
    min(train$life_expectancy),
    min(validate$life_expectancy),
    min(final_test_set$life_expectancy)
  ),
  Median = c(
    median(train$life_expectancy),
    median(validate$life_expectancy),
    median(final_test_set$life_expectancy)
  ),
  Max = c(
    max(train$life_expectancy),
    max(validate$life_expectancy),
    max(final_test_set$life_expectancy)
  )
)

summary_table[-1] <- round(summary_table[-1], 2)

kable(
  summary_table,
  caption = "Descriptive statistics of life expectancy across the data splits"
)
```

# EDA

If we examine global life expectancy over time (Figure 1), it is evident that 
life expectancy has increased substantially over the past decades. Overall, 
this suggests that global health outcomes have improved considerably.

```{r fig1, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

################################################################################
# EDA
################################################################################

#life expectancy world over the years

life_world <- life_at_birth %>%
  filter(entity == "World")

# fig 1.

ggplot(life_world,
       aes(x = year,
           y = life_expectancy,
           color = entity)) +
  geom_line(size = 1) +
  labs(
    title = "Fig 1. Life expectancy over time",
    x = "Year",
    y = "Life expectancy in years",
    color = ""
  ) +
  theme_minimal()
```

However, when comparing countries with the highest and lowest GDP per capita (Figure 2),
a clear disparity becomes visible. The poorest country in the dataset has a 
life expectancy below 60 years, whereas the richest country has a life expectancy 
exceeding 80 years. This represents a difference of more than 20 years. Furthermore,
when compared with the global average life expectancy of over 70 years, 
the poorest country falls approximately 10 years below the worldwide average.

```{r fig2, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

# find country with highest and lowest life expectancy

############################################################################
# Find the richest country using the most recent data point for each entity

gdp_current_rich <- gdp_per_capita %>%
  group_by(entity) %>%
  filter(year == max(year)) %>%  # Gets the latest year available for each country
  ungroup() %>%
  arrange(desc(gdp_pc)) %>%
  slice(1)

# Find the poorest country using the most recent data point for each entity

gdp_current_poor <- gdp_per_capita %>%
  group_by(entity) %>%
  filter(year == max(year)) %>%  # Gets the latest year available for each country
  ungroup() %>%
  arrange(gdp_pc) %>%
  slice(1)

# create a table

gdp_extremes <- bind_rows(
  gdp_current_rich %>%
    mutate(Category = "Richest Country"),
  gdp_current_poor %>%
    mutate(Category = "Poorest Country")
) %>%
  select(Category, entity, year, gdp_pc)

kable(
  gdp_extremes,
  col.names = c("Category", "Country", "Year", "GDP per Capita"),
  caption = "Countries with the Highest and Lowest GDP per Capita"
)

############################################################################

# plot life expectancy world, highest GDP and lowest GDP

data_plot <- life_at_birth %>%
  filter(entity %in% c("World", "Ireland", "South Sudan"),
         year >= 1950)

# fig 2.

ggplot(data_plot,
       aes(x = year,
           y = life_expectancy,
           color = entity)) +
  geom_line(size = 1) +
  labs(
    title = "Fig 2. life expectancy poor versus rich",
    x = "Year",
    y = "Life expectancy in years",
    color = "Country"
  ) +
  theme_minimal()
```

If we look at the target variable (life expectancy) (Figure 3), 
it becomes clear that the distribution is right-skewed. This indicates substantial 
variation across observations and suggests that many countries cluster around 
relatively high life expectancy values while a smaller number of countries have 
considerably lower values.

```{r fig3, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

################################################################################
# fig3. Target variable

ggplot(work_set, aes(x = life_expectancy)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 30) +
  labs(
    title = "Fig 3. Distribution of target variable",
    x = "Life expectancy in years",
    y = "Frequency"
  ) +
  theme_minimal()
```

A correlation matrix (Figure 4) reveals that the target variable is strongly 
correlated with several predictors, most notably GDP per capita, suggesting 
that economic conditions may be an important predictor of life expectancy and 
that GDP per capita itself contains substantial information about the target variable.

```{r fig4, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
################################################################################
# fig 4. correlation matrix

cor_matrix <- cor(
  work_set[, c(
    "life_expectancy",
    "spending_healthcare",
    "spending_edu",
    "years_in_school",
    "coverage__antigen_mcv1",
    "coverage__antigen_dtpcv3",
    "coverage__antigen_hib3",
    "coverage__antigen_rcv1",
    "share_obesity",
    "children_per_woman",
    "move_from_city",
    "move_to_city",
    "renewables_share_pct",
    "elect_dem_index",
    "gdp_pc"
  )],
  use = "complete.obs"
)

corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  title = "fig 4. Correlation predictors",
  mar = c(0, 0, 2, 0)
)
```

However, GDP per capita is also strongly correlated with several other predictors ,
suggesting that it may capture broader socio-economic conditions reflected in 
multiple indicators. Figure 5 shows that, after standardization, the distribution
of GDP per capita overlaps substantially with the distributions of many predictors. 
While this overlap does not imply redundancy on its own, it suggests that GDP 
per capita may share information with other variables in the dataset.

For instance, healthcare spending is strongly correlated with GDP per capita.
If we look at the distribution of healthcare spending and GDP per capita, it is 
noticeable that GDP covers a big area of healthcare spending, witch could indicated 
that some predictive information may already been captioned by other predictors in the dataset.

 Therefore, we hypothesize that the incremental predictive value of GDP per capita 
 may be smaller than expected once the remaining predictors are included in the model. 
 This hypothesis will be evaluated in the modelling section by comparing model 
 performance and variable importance measures.

```{r fig5, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
################################################################################
# fig 5. individual predictors with GDP as overlap

# scale
panel_scaled <- work_set %>%
  mutate(
    across(
      where(is.numeric),
      ~ as.numeric(scale(.))
    )
  )

# pivot longer
panel_long_scaled <- panel_scaled %>%
  select(-life_expectancy, -gdp_pc, -year, -entity) %>%
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "value"
  )

# fig 5.
ggplot(panel_long_scaled, aes(x = value)) +
  geom_density(fill = "skyblue", alpha = 0.5) +
  geom_density(
    data = panel_scaled,
    aes(x = gdp_pc),
    colour = "red",
    linewidth = 1,
    fill = NA,
    inherit.aes = FALSE
  ) +
  facet_wrap(
    ~ variable,
    scales = "free"
  ) +
  labs(
    title = " fig 5. Distribution of predictors with GDP per capita as overlay",
    x = "Value",
    y = "Density"
  ) +
  theme_minimal()
```

# Modelling

Before building the predictive models, it is important to consider which variables 
should be included in the analysis. The objective of this study is to identify 
socio-economic factors that are associated with life expectancy at birth. 

Variables such as the country itself contribute little to under-standing these 
underlying relationships. If a model learns that a particular country tends to 
have a higher or lower life expectancy, country-specific characteristics may 
obscure the effects of the actual predictors of interest. Therefore, the country 
variable was removed from the dataset prior to model training.

In addition, the year variable was excluded from the final models. 
The purpose of this study is to identify socio-economic factors associated with 
life expectancy, rather than to capture temporal trends. The year itself is not 
a direct explanatory factor but rather reflects broader societal, medical, and 
technological developments that have occurred over time. Excluding year allows 
the interpretation of the remaining predictors to remain more closely aligned 
with the research objective. Furthermore, it prevents the models from deriving 
a substantial portion of their predictive performance from the general upward 
trend in life expectancy over time, rather than from the socio-economic factors 
under investigation.

The modelling process begins with a simple baseline model, followed by a linear 
regression model, a k-nearest neighbours (KNN) model, and a Random Forest model. 
Model performance is evaluated using the Root Mean Squared Error (RMSE) metric.

The RMSE is defined as:

$$
RMSE =
\sqrt{
\frac{1}{n}
\sum_{i=1}^{n}
(\hat{y}-y)^2
}
$$

where:



- $y$ Actual life expectancy
- $\hat{y}$ Predicted life expectancy
- $n$ Number of observations



The RMSE measures the average magnitude of the prediction errors, giving greater weight to large errors due to the squaring operation. A lower RMSE indicates a more accurate predictive model.

To assess the predictive contribution of GDP per capita, the modelling procedure 
is conducted twice: once excluding GDP per capita and once including it as a 
predictor. 
As suggested in the exploratory data analysis, GDP per capita may 
share much of its explanatory power with other socio-economic variables.
Comparing model performance under both scenarios will therefore provide insight 
into the unique predictive value of GDP per capita.

Finally, the best-performing model, as determined by the validation RMSE, 
will be evaluated on the independent final test set to obtain an unbiased 
estimate of its predictive performance.

A simple baseline model yields an RMSE of approximately 7.9 years.
This result is based solely on using the mean life expectancy as the prediction.

```{r base_line, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

################################################################################
# modeling
################################################################################

################################################################################
# Simple base line
################################################################################

mean_life <- mean(train$life_expectancy)

pred_base <- rep(mean_life, nrow(validate))

rmse_simple <- sqrt(
  mean((pred_base - validate$life_expectancy)^2)
)

rmse_simple # 7.9
```

The linear model without GDP per capita already performs considerably better, 
reducing the RMSE to 4.5 years. Figure 6 shows the relationship between actual 
and predicted life expectancy. An interesting question is which variables 
contribute most to the model's performance. Fertility, measured as the number 
of children per woman, is by far the most important predictor.

```{r linear, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

################################################################################
# linear model
################################################################################

model <- lm(
  life_expectancy ~ spending_healthcare + spending_edu +
    years_in_school + coverage__antigen_mcv1 + coverage__antigen_dtpcv3 + 
    coverage__antigen_hib3 + coverage__antigen_rcv1 + share_obesity + children_per_woman +
    move_from_city + move_to_city + renewables_share_pct + elect_dem_index,
  data = train
)

pred <- predict(model, newdata = validate)

rmse_lm <- sqrt(
  mean((pred - validate$life_expectancy)^2)
)

rmse_lm # 4.5

# plot model

plot(validate$life_expectancy, pred,
     main = "Fig 6.: Actual vs predicted (linear)",
     xlab = "Actual life expectancy",
     ylab = "Predicted life expectancy",
     pch = 19,
     col = "steelblue")

abline(a = 0, b = 1, col = "red", lwd = 2)

tidy(model) %>%
  kable(digits = 3, caption = "Regression Model Results")

```

When the same model is estimated with GDP per capita included, the RMSE changes 
very little. Fertility remains the most important variable, 
indicating that GDP per capita adds limited predictive value beyond the 
information already captured by the other variables.

```{r linear_with_gdp_pc, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

################################################################################
# linear model with gdp_pc
################################################################################

model_gdp_pc <- lm(
  life_expectancy ~ spending_healthcare + spending_edu +
    years_in_school + coverage__antigen_mcv1 + coverage__antigen_dtpcv3 + 
    coverage__antigen_hib3 + coverage__antigen_rcv1 + share_obesity + children_per_woman +
    move_from_city + move_to_city + renewables_share_pct + elect_dem_index + gdp_pc,
  data = train
)

pred <- predict(model_gdp_pc, newdata = validate)

rmse_lm_gdp <- sqrt(
  mean((pred - validate$life_expectancy)^2)
)

rmse_lm_gdp # 4.5

tidy(model_gdp_pc) %>%
  kable(digits = 3, caption = "Regression Model Results with GDP")

```

Using the same set of predictors in a K-nearest neighbors (KNN) model,
but excluding GDP per capita, results in an RMSE of 2.3 years. Figure 7 illustrates 
the relationship between actual and predicted life expectancy. Compared with the 
linear model, the KNN model performs more accurately, suggesting that the relationship 
between the predictors and life expectancy is captured more effectively by a 
KNN model. In the KNN model, healthcare expenditure appears to be the 
most important predictor.


```{r knn, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

################################################################################
# KNN model
################################################################################

set.seed(123)

# train KNN
knn_model <- train(
  life_expectancy ~ spending_healthcare + spending_edu +
    years_in_school + coverage__antigen_mcv1 + coverage__antigen_dtpcv3 + 
    coverage__antigen_hib3 + coverage__antigen_rcv1 + share_obesity + children_per_woman +
    move_from_city + move_to_city + renewables_share_pct + elect_dem_index,
  data = train,
  method = "knn",
  preProcess = c("center", "scale"),
  tuneLength = 10
)

knn_model$bestTune

plot(knn_model)

# predict
pred_knn <- predict(knn_model, newdata = validate)

# RMSE
rmse_knn <- sqrt(
  mean((pred_knn - validate$life_expectancy)^2)
)

rmse_knn # 2.3

# plot
plot(
  validate$life_expectancy,
  pred_knn,
  main = "Figure 7: Actual vs predicted (KNN)",
  xlab = "Actual life expectancy",
  ylab = "Predicted life expectancy",
  pch = 19,
  col = "steelblue"
)

abline(a = 0, b = 1, col = "red", lwd = 2)

# variable importance

varimp_tbl <- varImp(knn_model)$importance %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Variable") %>%
  arrange(desc(Overall))

kable(
  varimp_tbl,
  digits = 3,
  caption = "Table 6. Variable importance for the KNN model"
)

```

When GDP per capita is added to the KNN model, the RMSE shows little to no improvement. 
However, GDP per capita exhibits greater predictive importance in this 
specification than in the linear model. Despite its increased contribution, 
the overall predictive performance of the model remains largely unchanged.

```{r knn_gdp, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

################################################################################
# KNN model with gdp_pc
################################################################################

set.seed(123)

# train KNN
knn_model_gdp <- train(
  life_expectancy ~ spending_healthcare + spending_edu +
    years_in_school + coverage__antigen_mcv1 + coverage__antigen_dtpcv3 + 
    coverage__antigen_hib3 + coverage__antigen_rcv1 + share_obesity + children_per_woman +
    move_from_city + move_to_city + renewables_share_pct + elect_dem_index + gdp_pc,
  data = train,
  method = "knn",
  preProcess = c("center", "scale"),
  tuneLength = 10
)

knn_model_gdp$bestTune

plot(knn_model_gdp)

# voorspellingen
pred_knn <- predict(knn_model_gdp, newdata = validate)

# RMSE
rmse_knn_gdp <- sqrt(
  mean((pred_knn - validate$life_expectancy)^2)
)

rmse_knn_gdp # 2.2

# variable importance

varimp_tbl <- varImp(knn_model_gdp)$importance %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Variable") %>%
  arrange(desc(Overall))

kable(
  varimp_tbl,
  digits = 3,
  caption = "Table 7. Variable importance for the KNN model with GDP"
)

```

If we apply the same procedure using a Random Forest model,
the resulting RMSE is 2 years. This performance is slightly better than that 
of the KNN model.

Figure 8 illustrates the relationship between actual and predicted life expectancy. 

In the Random Forest model, healthcare 
expenditure and fertility rate emerge as the most important predictors.

```{r RF, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

################################################################################
# rf model
################################################################################

set.seed(123)

tunegrid <- expand.grid(mtry = c(2,4,6,8,10))

rf_model <- train(
  life_expectancy ~ spending_healthcare + spending_edu +
    years_in_school + coverage__antigen_mcv1 + coverage__antigen_dtpcv3 + 
    coverage__antigen_hib3 + coverage__antigen_rcv1 + share_obesity + children_per_woman +
    move_from_city + move_to_city + renewables_share_pct + elect_dem_index,
  data = train,
  method = "rf",
  trControl = trainControl(method = "cv", number = 5),
  tuneGrid = tunegrid,
  ntree = 500
)

pred_rf <- predict(rf_model, newdata = validate)

rmse_rf <- sqrt(
  mean((pred_rf - validate$life_expectancy)^2)
)

rmse_rf # 2

plot(validate$life_expectancy, pred_rf,
     main = "Figure 8: Actual vs predicted (RF)",
     xlab = "Actual life expectancy",
     ylab = "Predicted life expectancy",
     pch = 19,
     col = "steelblue")

abline(a = 0, b = 1, col = "red", lwd = 2)

# variable importance

varimp_tbl <- varImp(rf_model)$importance %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Variable") %>%
  arrange(desc(Overall))

kable(
  varimp_tbl,
  digits = 3,
  caption = "Table 8. Variable importance for the RF model"
)

plot(varImp(rf_model))

```

When GDP per capita is added to the Random Forest model, the RMSE does not
improve that much. It settles around 1.98 years. Similar to the KNN results, 
GDP per capita appears to have a 
higher variable importance within the model, but it's inclusion does not lead 
to a meaningful improvement in predictive performance. This finding suggests 
that much of the information contained in GDP per capita is already captured 
by the other socio-economic indicators included in the analysis.

```{r RF_gdp, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

################################################################################
# rf model with gdp_pc
################################################################################

set.seed(123)

tunegrid <- expand.grid(mtry = c(2,4,6,8,10))

rf_model_gdp <- train(
  life_expectancy ~ spending_healthcare + spending_edu +
    years_in_school + coverage__antigen_mcv1 + coverage__antigen_dtpcv3 + 
    coverage__antigen_hib3 + coverage__antigen_rcv1 + share_obesity + children_per_woman +
    move_from_city + move_to_city + renewables_share_pct + elect_dem_index + gdp_pc,
  data = train,
  method = "rf",
  trControl = trainControl(method = "cv", number = 5),
  tuneGrid = tunegrid,
  ntree = 500
)

pred_rf <- predict(rf_model_gdp, newdata = validate)

rmse_rf_gdp <- sqrt(
  mean((pred_rf - validate$life_expectancy)^2)
)

rmse_rf_gdp # 1.98

# variable importance

varimp_tbl <- varImp(rf_model_gdp)$importance %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Variable") %>%
  arrange(desc(Overall))

kable(
  varimp_tbl,
  digits = 3,
  caption = "Table 9. Variable importance for the RF model with GDP"
)

plot(varImp(rf_model_gdp))

```

Variable Importance. Based on the variable importance measures of the
best performing model, the Random Forest model including GDP per capita, 
it became evident that the variable share of obesity contributed little to no 
predictive value. As its importance was effectively zero, this variable was 
excluded from the final model.

Although GDP per capita did not substantially improve predictive performance, 
the Random Forest model including GDP per capita was selected for evaluation 
on the independent final test set. This model achieved an RMSE of 1.8 years, 
indicating that life expectancy can be predicted with a relatively high degree 
of accuracy using the selected socio-economic indicators.

```{r final, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
################################################################################
# test RF model with GDP on final_test_set
################################################################################

set.seed(123)

tunegrid <- expand.grid(mtry = c(2,4,6,8,10))

rf_model_best <- train(
  life_expectancy ~ spending_healthcare + spending_edu +
    years_in_school + coverage__antigen_mcv1 + coverage__antigen_dtpcv3 + 
    coverage__antigen_hib3 + coverage__antigen_rcv1 + children_per_woman +
    move_from_city + move_to_city + renewables_share_pct + elect_dem_index + gdp_pc,
  data = train,
  method = "rf",
  trControl = trainControl(method = "cv", number = 5),
  tuneGrid = tunegrid,
  ntree = 500
)

pred_rf <- predict(rf_model_best, newdata = final_test_set)

rmse_final_test <- sqrt(
  mean((pred_rf - final_test_set$life_expectancy)^2)
)

rmse_final_test # 1.8
```

Table 9 summarizes the predictive performance of all models evaluated on the validation set. 
The results show a substantial improvement over the baseline model for all 
machine learning approaches. The Random Forest model achieved the lowest validation 
RMSE and was therefore selected for final evaluation on the independent test set. (Table 10)

```{r summary, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
################################################################################
# Summary
################################################################################

model_results <- data.frame(
  Model = c(
    "Baseline",
    "Linear Regression",
    "Linear Regression + GDP",
    "KNN",
    "KNN + GDP",
    "Random Forest",
    "Random Forest + GDP"
  ),
  Validation_RMSE = c(
    rmse_simple,
    rmse_lm,
    rmse_lm_gdp,
    rmse_knn,
    rmse_knn_gdp,
    rmse_rf,
    rmse_rf_gdp
  )
)

model_results$Validation_RMSE <- round(
  model_results$Validation_RMSE,
  2
)

kable(
  model_results,
  caption = "Table 10. Validation RMSE for all evaluated models"
)

final_results <- data.frame(
  Model = "Random Forest + GDP",
  Validation_RMSE = round(rmse_rf_gdp, 2),
  Final_Test_RMSE = round(rmse_final_test, 2)
)

kable(
  final_results,
  caption = "Table 11. Performance of the final selected model"
)
```

Although only a limited number of predictors are included in this analysis, 
a model such as this could still provide useful insights for policymakers seeking 
to improve population health outcomes and increase life expectancy. 
Additional health, environmental, and social indicators could also be incorporated 
to determine whether prediction accuracy can be improved further.

# Conclusion

This study examined whether life expectancy at birth can be predicted using a
range of socio-economic indicators and investigated which factors contribute 
most strongly to these predictions.
Referring to the first and third research questions, the results show that life 
expectancy can be predicted with reasonable accuracy using socio-economic indicators.
The best-performing model was a RF model, which achieved a 
final test RMSE of 1.8 years. This indicates that life expectancy can be 
predicted within approximately two years on average, representing a substantial 
improvement over the baseline model.
With respect to the second research question, healthcare expenditure and fertility rate
emerged as the most influential predictors across several modelling approaches. 
Education, vaccination cover-age, and other socio-economic indicators also 
contributed to the predictive performance of the models.
An interesting finding is that GDP per capita, despite being strongly associated 
with life expectancy, added relatively little predictive value once other 
socio-economic indicators were included in the analysis. This suggests that
the effects of national wealth may already be reflected through factors such as 
healthcare, education, and other investments that directly influence population health.

Several limitations should be acknowledged. The analysis was restricted to the 
variables available from the selected data sources, and substantial missing data 
required imputation. Furthermore, the results represent statistical associations 
and should not be interpreted as causal relationships.
Overall, the findings suggest that socio-economic indicators can provide 
meaningful insights into differences in life expectancy. While economic prosperity 
remains important, the way resources are translated into healthcare, education, 
and broader social development appears to be at least as important as income levels alone.


# References

World Health Organization (n.d.) Health inequality monitoring: harnessing data to advance health equity. Available at: https://iris.who.int/server/api/core/bitstreams/7ed039ab-4a7f-4f66-9b4b-264008d98f81/content (Accessed: 20 September 2026).

Our World in Data (n.d.) Our World in Data. Available at:
https://ourworldindata.org/ (Accessed: 20 September 2026).

Stekhoven, D.J. and Bühlmann, P. (2012) ‘MissForest: non-parametric missing value 
imputation for mixed-type data’, Bioinformatics, 28(1), pp. 112–118. Available at:
https://doi.org/10.1093/bioinformatics/btr597.

Use of Artificial Intelligence

Microsoft Copilot was used to assist with spelling and grammar checks and to
provide suggestions for structuring the project and identifying potential analysis 
approaches. All data collection, coding, modelling, interpretation of results, and 
final writing decisions were performed by the author.
